Natural Selection Across Three Generations of Americans

### 2024-02-16 response-to-reviewers-revision-1.qmd

The earlier claim is not "coefficients are negative overall" but 
"coefficients among younger parents are negatively correlated with coefficients
among older parents". We are dealing with correlations of correlations: this
can get confusing.

Negatively correlated, and more strongly, opposite-signed, effects between
young and old parents support the theory that income effects dominate for
older parents while substitution effects dominate for younger parents. To
be a bit more explicit: if a PGS simply doesn't correlate with human capital,
we expect it to not be selected on among either group of parents. If a PGS
is correlated with human capital, that will lead to income and substitution effects
for those with more human capital. If the substitution effect dominates for
younger parents then the PGS will have a negative correlation with fertility;
if the income effect dominates for older parents, then the PGS will have a positive
correlation with fertility.

The test against 29 had the following logic: if all effects were really zero,
and estimates were pure noise with mean 0, then we'd expect opposite-signed and same-signed "effects" to be equally likely; so, 29 is half of the 58 scores. However,
this is a rather specific null. A more natural null is that selection effects
are non-zero, but the same among both groups of parents. Then 
young-parent effects would be equally likely to be higher, or lower, than old-parent
effects.

DHJ TODO: can we figure out a test for this?

* Test: take difference between scores, times the sign of the low group score.
  Then under the null we'd expect as many + as - differences. Do this and compare
  to 29, always, put it in table 1.

@@

We have clarified the theory early on in the intro and abstract.
TE TODO: Write a paragraph on biol. selection. (Not sure what is meant here?)

In terms of accepting or rejecting our theory, our paragraph in the discussion
was:

> The results here support prediction 1 but are more ambiguous for prediction 2. PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents. But there is little evidence for larger coefficients among people with lower education, or younger parents. This may be due to the low sample size. But in the UK, the between-group differences were large; differences that big would surely have been visible here. The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason. But note that *any* theory with a negative relationship between education and fertility will give prediction 1. In this sense, results here are less supportive of the economic theory specifically.

We don't think these results give a simple thumbs up or down to the theory,
but we're open to suggestions that would help clarify what we say.

@@

*What is the data base for the mean number of children of people born in the same year - is pooling appropriate, as the means may vary considerably according to ethnicity?*

Means do vary (TODO how much by?). But note that we analyse the ethnicities
separately anyway, so the only effect of pooling is to scale all the selection effects
within each ethnicity.

As a matter of scientific correctness, we think it's mostly best to think of both
ethnic groups as one population. Interracial marriage was very low for much
of the twentieth century, but this was a fact about laws and social norms, not
about biology.

DHJ/TE TODO: change it to be separate RLRS baseline for white and black; add 
a couple of sentences discussing the implications. (Effectively examining a
subpopulation as an imperfect proxy for the whole population.)

### 2024-02-19 hrs-selection.qmd
Put black sample in appendix. Rework statistical tests and graphs.

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.

@@

We estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes. In the white sample, `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, and self-rated health. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

```{r}
#| label: stat-bootstraps
#| cache: false
#| cache-lazy: false
#| cache-comments: false

@@

@@

@fig-rlrs-edu plots each PGS's regression beta on RLRS against its regression beta on educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   Controls include 10 principal components of genetic array
#|   data. Dashed lines are fitted from linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

for (eth in eths) {
  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
          geom_hline(yintercept = 0, color = "grey80") +
          geom_vline(xintercept = 0, color = "grey80") +
          geom_point() +
          geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                      linetype = "dashed", colour = "red", linewidth = 0.5) +
          scale_y_continuous(breaks = seq(-0.2, 0.2, 0.02),
                             labels = scales::label_number(0.01)) +
          coord_cartesian(xlim = c(-.3, .75)) +
          theme_light() + 
          theme(panel.grid = element_blank()) +
          labs(
            title = titles[eth],
            x = "Coefficient on years of education",
            y = "Coefficient on RLRS"
          )
  print(ggp)
}

@@

```{r}
#| label: stat-siblings

tmp <- 
  tidy_pgs_sibs$black |> 
  inner_join(tidy_pgs_ed$black, 
             by = join_by(term), 
             suffix = c(".sibs", ".edyrs")) |> 
  filter(term != "GWAD2NA", term != "GWAD2WA")

cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
# if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
# TE correlation without outliers is not NEG
# But it is not negative anyway now
```

```{r}
#| label: stat-bootstrap-gen-cor
#| cache: false

@@

```{r}
#| label: stat-bootstrap-gen-cor-gkids
#| cache: false

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are only significant and negative in the white respondents, although ethnicities have large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

To examine selection in the succeeding generation, we divide the respondents' number of grandchildren by their number of children. In other words, we calculate the average number of children the respondent's children have. This indicates reproductive success in the second generation which we can call RLRS2. `r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having reported no children. These individuals were removed for this test. In regression we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.

Older grandparents will have more time for their number of grandkids to accumulate. To better deal with this time trend, we subset the data to parents whose oldest child was $40$ years old or older in $2016$, when the number of grandchildren was recorded. Appendix @fig-offspring-cohorts shows the distribution of the birth year of the oldest child, adjusted for sampling weights. A majority of the observations are after $1960$, the most recent year the respondents were born in. This reassures us that we are observing reproduction in later generations.

We regress the respondent's (the grandparent's) PGS on RLRS2. The resulting coefficient is only a proxy for selection in the next generation, not a perfect measure. The grandparent's PGS indicates, but does not determine the PGS of the parent. To know the expected parent's PGS, we would need both grandparent's PGS. Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, such an approach is untenable in our sample. We expect the effect of the grandparent's PGS to also be dependent upon the level of assortative mating for the trait. For traits with high assortative mating, the grandparent's PGS will more strongly predict the parent's PGS, leading to a greater regression slope.

Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education. Correlations are significant for whites but not blacks. Standard errors were large in both ethnicites (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`; blacks: correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`). There is a positive correlation between PGS coefficients on RLRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I. `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.

We next test part 2 of the theory by interacting PGS with measures of education, income, marital status, and age at first birth. Education is years of education, split at 12 years. Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split. From here on we only use the white subsample: there are too few black respondents to be informative.

```{r}
#| label: stat-x-bootstrap-cors
#| cache: false
#| cache-lazy: false
#| eval: false

old_opts <- options(survey.multicore = TRUE)

# all confindence intervals were too wide to be informative
n_reps_x <-  199

calc_diff_cor_ed <- function (w, data, term) {
  groups <- switch(term,
                edu = c("0-12", "13-17"),
                income_med = c("Low", "High"),
                married = c("Other", "Married"),
                agefb = c("Low", "High"))
  groups <- paste0(":", term, groups) # colon for the cross term
  low_group <- groups[1]
  high_group <- groups[2]

  forms <- paste0(pgs, ":", term) |>
    map(\(x) reformulate(c(x, term, pcs), response = "rlrs"))
  mods <- map(forms, \(x) lm(x, data, weights = w))
  cfs <- map(mods, coef)
  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])

  forms_ed <- paste0(pgs, ":", term)  |> 
    map(\(x) reformulate(c(x, term, pcs), response = "raedyrs"))
  mods_ed <- forms_ed |> map(\(x) lm(x, data, weights = w))
  cfs_ed <- map(mods_ed, coef)
  cfs_ed_low <- map_dbl(cfs_ed, \(x) x[str_detect(names(x), low_group)])
  cfs_ed_high <- map_dbl(cfs_ed, \(x) x[str_detect(names(x), high_group)])

  cor_ed_low <- cor(cfs_low, cfs_ed_low)
  cor_ed_high <- cor(cfs_high, cfs_ed_high)

  c(low = cor_ed_low, high = cor_ed_high, diff = cor_ed_low - cor_ed_high)
}

diff_cor_ed <- ci <- est <- tbl <- list()

terms <- c("edu", "income_med", "married", "agefb")
for (eth in eths) {
  rand_boot <- rand |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps_x)

  for (term in terms) {
    diff_cor_ed[[eth]][[term]] <- 
      withReplicates(rand_boot, theta = calc_diff_cor_ed, term = term)  
    ci[[eth]][[term]] <- confint(diff_cor_ed[[eth]][[term]])
    est[[eth]][[term]] <- as.numeric(diff_cor_ed[[eth]][[term]])
  }
}

options(old_opts)

@@

```{r}
#| label: stat-x-bootstrap-effect-size
#| cache: false
#| cache-lazy: false
#| cache-comments: false

n_reps_x_es <-  199
pgs <- pgs_white

calc_diff_effect_size <- function (w, data, term) {
  groups <- switch(term,
                edu = c("0-12", "13-17"),
                income_med = c("Low", "High"),
                married = c("Other", "Married"),
                agefb = c("Low", "High"),
                born = c("pre-1942", "post-1942"))
  groups <- paste0(":", term, groups) # colon for the cross term
  low_group <- groups[1]
  high_group <- groups[2]

  forms <- paste0(pgs, ":", term) |>
    map(\(x) reformulate(c(x, term, pcs), response = "rlrs"))
  mods <- map(forms, \(x) lm(x, data, weights = w))
  cfs <- map(mods, coef)
  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])

  abs_eff_low <- mean(abs(cfs_low))
  abs_eff_high <- mean(abs(cfs_high))
  diff_abs_eff <- abs_eff_low - abs_eff_high
  c(low = abs_eff_low, high = abs_eff_high, diff = diff_abs_eff)
}

effect_sizes <- ci <- est <- tbl <- list()

for (term in terms) {
  effect_sizes[[term]] <- 
    withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
  ci[[term]] <- confint(effect_sizes[[term]])
  est[[term]] <- as.numeric(effect_sizes[[term]])
  tbl[[term]] <- sprintf("%.3f (%.3f to %.3f)", est[[term]], 
                         ci[[term]][, 1], ci[[term]][, 2])
}

```{r}
#| label: stat-bootstrap-agefb
#| cache: false

n_reps_agefb <-  199
pgs <- pgs_white

calc_cor_agefb <- function (w, data) {
  forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))
  mods_low <- map(forms, \(x) lm(x, data, subset = agefb == "Low",
                                 weights = w))
  mods_high <- map(forms, \(x) lm(x, data, subset = agefb == "High",
                                  weights = w))
  coefs_low <- map2_dbl(mods_low, pgs, \(x, y) coef(x)[[y]])
  coefs_high <- map2_dbl(mods_high, pgs, \(x, y) coef(x)[[y]])

  c(
    cor = cor(coefs_low, coefs_high), 
    discordant = sum(sign(coefs_low) != sign(coefs_high)) # discordant pairs
  )
}

rand_boot <- rand |> 
  subset(ethnicity == "white") |> 
  as.svrepdesign(type = "bootstrap", replicates = n_reps_agefb)

cor_agefb <- withReplicates(rand_boot, theta = calc_cor_agefb)

stopifnot(insignificant(confint(cor_agefb)[1,]))
ci_discordant <- confint(cor_agefb)[2,]
# significantly different from 1/2?
stopifnot(ci_discordant[1] <= length(pgs)/2 && 
            ci_discordant[2] >= length(pgs)/2)
# TE THEY ARE DISCORDANT

```{r}
#| label: stat-post-x-bootstrap

stopifnot(insignificant(ci$edu[3,]))
stopifnot(significant(ci$income_med[3,]))
stopifnot(significant(ci$married[3,]))
stopifnot(insignificant(ci$agefb[3,]))

tidy_agefb_wide <- tidy_pgs_x_agefb$white |> 
  pivot_wider(id_cols = pgs, names_from = agefb, values_from = estimate)

cor_agefb <- cor(tidy_agefb_wide$Low, tidy_agefb_wide$High)
n_discordant <- sum(sign(tidy_agefb_wide$Low) != sign(tidy_agefb_wide$High))
```

Appendix @fig-x-edu shows essentially no difference in the distribution of absolute coefficient sizes between respondents with more and less education. Appendix @fig-x-inc, on the other hand, shows that coefficients appear larger among lower-income respondents. Appendix @fig-x-mar shows absolute coefficients for respondents who were married in 2010 against all other statuses. Coefficients appear larger for unmarried respondents.

Appendix @fig-x-agefb splits respondents by age at first birth (median within each gender). Note that the N is lower here due to missing values. There are no large differences in absolute effect size between younger and older parents. Just as in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs. However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively. Also, in the UK Biobank sample, PGS coefficients on fertility controlling for age at first birth were negatively correlated with the uncontrolled coefficients; here, that isn't true, with coefficients remaining positively correlated.

@@

@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in each pair of groups, and the difference between them, using bootstraps. Differences for income and marriage are significant.

|               | Low group             | High group            | Difference            |
|------------------|------------------|------------------|------------------|
| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
| Birth year    | `r tbl$born[1]`       | `r tbl$born[2]`       | `r tbl$born[3]`       |

: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth; born before vs. after 1942. `r n_reps_x_es` bootstraps. {#tbl-x-bootstraps tbl-colwidths="\[19,28,28,33\]"}

The economic theory of fertility also implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^1] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups. we tested this with bootstraps, but confidence intervals were always too wide to be informative.

[^1]: See @hugh2022human equation (6) and following.

@@

SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository. The authors use GCTA to estimate heritability. When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead. We also perform error correction with twin heritabilites. To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^2]. We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power. This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^3].

[^2]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin. For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].

>>>>

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans. To preview our results, we confirm point 1 above, but only see partial and ambiguous support for point 2. (TODO: or can we be stronger?)

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants. We focus on these because the sample size is large enough. The appendix reports the most basic analyses for the `r n_black` genotyped black participants.

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants. PGS are rescaled to zero mean and unit variance. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data. For our purposes, this is equivalent to residualizing on principal components [@wurm2014residualizing]. Effectively, our analysis is of the residualized PGS that are widely used in practical applications. \[TODO actually we could just residualize; only the SE of the PCs is affected\]

@@

We estimate coefficients of PGS on RLRS. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, and self-rated health. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

```{r}
#| label: stat-bootstraps
#| cache: true
#| cache-lazy: false
#| cache-comments: true

# TODO set cache to false before final build

@@

  pgs <- if (eth == "white") pgs_white else pgs_black

@@

@fig-rlrs-edu plots each PGS's regression beta on RLRS against its regression beta on educational attainment. The relationship is negative (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   Controls include 10 principal components of genetic array
#|   data. Dashed lines are fitted from linear regressions.

ggplot(tidy_pgs_joined$white, aes(estimate.edyrs, estimate.fert)) +
   geom_hline(yintercept = 0, color = "grey80") +
   geom_vline(xintercept = 0, color = "grey80") +
   geom_point() +
   geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
               linetype = "dashed", colour = "red", linewidth = 0.5) +
   scale_y_continuous(breaks = seq(-0.2, 0.2, 0.02),
                      labels = scales::label_number(0.01)) +
   coord_cartesian(xlim = c(-.3, .75)) +
   theme_light() + 
   theme(panel.grid = element_blank()) +
   labs(
     x = "Coefficient on years of education",
     y = "Coefficient on RLRS"
   )

@@

```{r}
#| label: stat-bootstrap-gen-cor
#| cache: true
#| cache-comments: true

# TODO set cache to FALSE before submission

@@

```{r}
#| label: stat-bootstrap-gen-cor-gkids
#| cache: true
#| cache-comments: true

# TODO set cache to FALSE before submission

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are significant and negative (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (`r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

To examine selection in the next generation, we divide the respondents' number of grandchildren by their number of children. In other words, we calculate the average number of children of the respondent's children. This indicates reproductive success in the second generation, which we call RLRS2. We removed `r gkids_unusable_white` respondents who report having grandchildren despite reporting having reported no children. In regressions we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.

Older grandparents will have more time for their number of grandkids to accumulate. To deal with this time trend, we subset the data to parents whose oldest child was $40$ years old or older in $2016$, when the number of grandchildren was recorded. Appendix @fig-offspring-cohorts shows the distribution of the birth year of the oldest child, adjusted for sampling weights. A majority of the observations are after $1960$, the most recent year the respondents were born in. This reassures us that we are observing reproduction in later generations.

We regress the respondent's (the grandparent's) PGS on RLRS2. The resulting coefficient is only a proxy for selection in the next generation, not a perfect measure. The grandparent's PGS indicates, but does not determine the PGS of the parent. To know the expected parent's PGS, we would need both grandparents' PGS. Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, that approach is untenable in our sample. We expect the effect of the grandparent's PGS to also depend upon the level of assortative mating for the trait. For traits with high assortative mating, the grandparent's PGS will more strongly predict the parent's PGS, leading to a greater regression slope.

Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education. Correlations are significant. Standard errors are large (correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`). There is a positive correlation between PGS coefficients on RLRS2 and RLRS (`r gen_cor_gkids`, bootstrapped 95% C.I. `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.

## Differences across social groups

We next test part 2 of the theory by interacting PGS with measures of education, income, marital status, and age at first birth. Education is years of education, split at 12 years. Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split. We call the low-education, low-income, unmarried or younger-AFB group the "disadvantaged group".

In @hugh2022human we could simply look for significant differences between coefficients on RLRS among the groups, for each polygenic score. That will not work here because of the smaller sample size. At the same time, scores are not independent and not a sample from a population. To get round this, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

$$
RLRS_i=\alpha+\beta PGS_i + \gamma DIS_i + \delta (PGS_i \times DIS_i) + \varepsilon_i
$$

where $DIS_i$ is a dummy for $i$'s membership in the disadvantaged group and where $PGS_i$ has been sign-flipped to correlate negatively with education. Under the null, the estimated $\delta$s will be distributed around zero; in expectation half will be positive and half negative. Under the alternative, more than half the $\delta$s will be positive. We bootstrap respondents and count the number of positive $\delta$s in each resample. If 95% confidence intervals are above `r length(pgs_white)`/2 = `r length(pgs_white)/2`, we reject the null in favour of the alternative. \[TODO: what if some are negative and some are positive? I think this is fine because it makes our test conservative? Maybe??\]

```{r}
#| label: stat-x-bootstrap-effect-size
#| cache: true
#| cache-lazy: false
#| cache-comments: true

# TODO set cache to false before submission

n_reps_x_es <-  199

calc_delta_coefs <- function (w, data, term) {
  groups <- switch(term,
                   edu = c("0-12", "13-17"),
                   income_med = c("Low", "High"),
                   married = c("Other", "Married"),
                   agefb = c("Low", "High"),
                   born = c("pre-1942", "post-1942"))
  groups <- paste0(":", term, groups) # colon for the cross term
  low_group <- groups[1]
  high_group <- groups[2]

  # We estimate separate betas then take the difference,
  # rather than explicitly using a cross term.
  # This is simpler.
  cross_terms <- paste0(pgs_white, ":", term)

  # "rlrs ~ pgs:term + term + pc1 + pc2 +..."
  forms <-  map(cross_terms, 
                \(x) reformulate(c(x, term, pcs), response = "rlrs"))
  mods <- map(forms, \(x) lm(x, data, weights = w))
  cfs <- map(mods, coef)
  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])

  # \delta in the estimating equation above.
  delta_coefs <- cfs_low - cfs_high

  setNames(delta_coefs, pgs_white)
}

calc_num_positive_deltas <- function (w, data, term) {
  delta_coefs <- calc_delta_coefs(w, data, term)
  # the name is required to workaround a bug in survey
  c(n_pos = sum(delta_coefs > 0))
}

# sign-flip the PGS to correlate negatively with education
for (pgs_name in pgs_white) {
  pgs <- rand_boot$variables[[pgs_name]]
  sign_cor_pgs_edu <- sign(cor(pgs, rand_boot$variables$raedyrs, 
                               use = "complete"))
  rand_boot$variables[[pgs_name]] <- pgs * -1 * sign_cor_pgs_edu
}

# to get the non-bootstrapped deltas do, e.g.:
# calc_delta_coefs(weights(rand_boot, type = "sampling"), 
#                  rand_boot$variables, "edu")

n_pos_deltas <- ci <- est <- tbl <- list()

for (term in terms) {
  n_pos_deltas[[term]] <- 
    withReplicates(rand_boot, theta = calc_num_positive_deltas, term = term)

  ci[[term]] <- confint(n_pos_deltas[[term]])
  est[[term]] <- round(n_pos_deltas[[term]])
  tbl[[term]] <- sprintf("%d (%.1f to %.1f)", est[[term]], 
                         ci[[term]][, 1], ci[[term]][, 2])
}

# calculate non-bootstrapped deltas
sampling_weights <- weights(rand_boot, type = "sampling")

deltas <- map(terms, \(term) {
                calc_delta_coefs(sampling_weights,
                                 rand_boot$variables, 
                                 term)
              })

deltas <- deltas |> 
  setNames(terms) |> 
  as.data.frame() |> 
  as_tibble(rownames = "PGS") |> 
  pivot_longer(-PGS, names_to = "group", values_to = "delta")

rm(rand_boot) # avoid messing with subsequent code

@@

```{r}
#| label: fig-deltas
#| fig-cap: Coefficients of interaction terms ($\delta$) for effects of
#|   polygenic scores 
#|   on RLRS among different social groups. A positive $\delta$ means that
#|   the coefficient was larger for the disadvantaged group, where the score
#|   was sign-flipped to correlate negatively with education. Horizontal lines
#|   show medians.
deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(rank(delta) %in% c(1:3, 66:68), PGS, ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, position = position_jitter(width = 0.1)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

Figure @fig-deltas plots values of $\delta$ for each interaction term, without bootstrapping. The majority of $\delta$s are positive for age at first birth, income and marriage but not for education. In other words, for younger parents, poorer participants, and unmarried participants, most PGS have larger effects on RLRS, but this is not true for less educated participants.

@@

@tbl-x-bootstraps shows estimates and bootstrapped 95% confidence intervals for the number of scores where the $\delta$ term is positive. Confidence intervals exclude 34 only for income.

|                    | Estimate (95% C.I.) |
|--------------------|---------------------|
| Education          | `r tbl$edu`         |
| Income             | `r tbl$income_med`  |
| Marriage           | `r tbl$married`     |
| Age at first birth | `r tbl$agefb`       |
| Birth year         | `r tbl$born`        |

: Estimates and bootstrap 95% confidence intervals for numbers of PGS out of `r length(pgs_white)` where the disadvantaged group had a larger selection coefficient than the advantaged group. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth; born before vs. after 1942. `r n_reps_x_es` bootstraps.

@@

SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository. The authors use GCTA to estimate heritability. When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead. We also perform error correction with twin heritabilites. To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^1]. We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power. This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2].

[^1]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin. For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].

[^2]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms. These were PV001-PV018 in the 2016 core survey. They estimated an $R^2_{PGS}$ of $0.003$. When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`. Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I. `r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`. Given the great uncertainty over the true correlation we are unable to justify using the error-corrected estimator with the ADHD polygenic score.
